---
title: "DATA624 - Homework 1"
author: "Esteban Aramayo"
date: "2022-06-05"
# output: openintro::lab_report
output: word_document
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE,
                      message=FALSE,
                      collapse = FALSE,
                      comment = "#>" )
```



# Week 1, Getting Started and Time Series Graphics [May 31 - June 05]



<br>

## Exercise 2.10.1 

Use the help function to explore what the series gold, woolyrnq and gas represent.

a. Use autoplot() to plot each of these in separate plots.


### Solution

<br>

```{r}
library(fpp2)
```

```{r}
autoplot(gold) +
  ggtitle("Daily morning gold prices in US dollars") +
  ylab("Gold Price") +
  xlab("Day")
```

```{r}
autoplot(woolyrnq) +
  ggtitle("Quarterly production of woollen yarn in Australia: tonnes") +
  ylab("Yarn Production") +
  xlab("Year-Qrt")
```

```{r}
autoplot(gas) +
  ggtitle("Australian monthly gas production") +
  ylab("Gas Production") +
  xlab("Year-Month")
```

<br>

b. What is the frequency of each series? Hint: apply the frequency() function.

### Solution

<br>

The frequency of the $gold$ series is annual because its frequency value is `r frequency(gold)`.

```{r}
frequency(gold)
```

The frequency of the $woolyrnq$ series is quarterly because its frequency value is `r frequency(woolyrnq)`.

```{r}
frequency(woolyrnq)
```

The frequency of the $gas$ series is monthly because its frequency value is `r frequency(gas)`.

```{r}
frequency(gas)
```



<br>

c. Use which.max() to spot the outlier in the gold series. Which observation was it?

### Solution

<br>

The outlier observation in the $gold$ series is `r gold[which.max(gold)]`, which is located at the position `r which.max(gold)`.

```{r}
gold[which.max(gold)]
```


<br>




## Exercise 2.10.3 

Download some monthly Australian retail data from the book website. These represent retail sales in various categories for different Australian states, and are stored in a MS-Excel file.

a. You can read the data into R with the following script:

retaildata <- readxl::read_excel("retail.xlsx", skip=1)

The second argument (skip=1) is required because the Excel sheet has two header rows.

### Solution

<br>

```{r}

# library(rio)
# retail_url <- "https://otexts.com/fpp2/extrafiles/retail.xlsx"
# retaildata <- rio::import(file = retail_url, which = 1)
# head(retaildata)

library(readxl)
retaildata <- readxl::read_xlsx("retail.xlsx", skip = 1)
head(retaildata)

```


<br>

b. Select one of the time series as follows (but replace the column name with your own chosen column):

myts <- ts(retaildata[,"A3349873A"], frequency=12, start=c(1982,4))

### Solution

<br>

```{r}
myts <- ts(retaildata[,"A3349336V"], frequency=12, start=c(1982,4))

myts
```


<br>

c. Explore your chosen retail time series using the following functions:

autoplot(), ggseasonplot(), ggsubseriesplot(), gglagplot(), ggAcf()

Can you spot any seasonality, cyclicity and trend? What do you learn about the series?



### Solution

